---
title: 'Assignment 03: Vectors and matrices'
author: 'Kenji Sato^[Kobe University. mail@kenjisato.jp]'
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  bookdown::pdf_document2:
    fig_caption: yes
    includes:
      in_header: asset/latex/custom.sty
    keep_tex: no
    latex_engine: pdflatex
    number_sections: yes
    template: null
    toc: no
  bookdown::html_document2: default
documentclass: scrartcl
---

# Overview

## Purpose {-}

To learn how to use vectors and matrices with R. Write user defined functions.

## Instructions {-}

In this assignment, you will

- clone the assignment repository and make a working branch (eg. `solution` branch);
- solve the problems in Section \@ref(problems); 
- write the solutions in `R/matrix.R`, edit and knit `solution.Rmd`; 
- commit `R/matrix.R`, `solution.Rmd` and `solution.pdf`; and 
- open a Pull Request.


# R quick course

## Atomic vectors

A numeric vector is a tuple of numbers. In mathematics, when we write such an expression as
$$
x = 
\begin{bmatrix}
  2.1 \\
  1.2 \\
  4.3
\end{bmatrix}
$$
we understand that this $x$ is a vector in the three-dimensional vector space (it can be $\mathbb R^3$, $\mathbb C^3$ or $\mathbb Q^3$). 

In R, we use `c()` function to concatenate the three numbers.
```{r}
x <- c(2.1, 1.2, 4.3)
x
```

An object like `x` is called an atomic vector. All elements have a common type and are aligned flat and straight. 

To read a value in the vector, we use the indexing operator `[`.^[This is just a R function. Try `` `r '\x60[\x60(x, 1)'` `` in the console.] Note that the index starts from 1.^[Notice that the $n$-th element of $x$ in this example has $n$ after the decimal point. I owe this idea to Hadley Wickham. See http://adv-r.had.co.nz/Subsetting.html.]

```{r}
x[1]
```

Similarly, 

```{r}
x[2]
```

We can use multiple subscripts  

```{r}
x[c(2, 3)]
```

and negative subscript

```{r}
x[-1]
```

Negative multiple subscripts works just fine but you cannot mix positive and negative subscripts.

```{r, error=TRUE}
x[c(1, -2)]
```

## Scalars

In R, there is no such thing as a scalar value. A single number is a vector of size one.

```{r}
1 == c(1)
```

The following code might seem a bit awkward at first glance. 

```{r}
length(0)
```


## Arithmetics

Addition and subtraction of two (or more) vectors with same size work as we expect. 

```{r}
y <- c(-2, -1, -4)
x + y
```

Multiplication and division are performed element wise.

```{r}
x * y
```

## Column and row vectors

In mathematics, 
$$
x = 
\begin{bmatrix}
  2.1 \\
  1.2 \\
  4.3
\end{bmatrix}
$$
and
$$
x = 
\begin{bmatrix}
  2.1 & 1.2 & 4.3
\end{bmatrix}
$$
are usually considered different. The first is called a column vector and the latter a row vector. `c(2.1, 1.2, 4.3)` doesn't have row-column distinction.

To get a column vector, use `matrix()` function
```{r}
xcol <- matrix(x)
xcol
```

To get a row vector, do the following
```{r}
xrow <- matrix(x, nrow = 1)
xrow
```

Alternatively, you can do something like this. `t()` function returns the transpose matrix.

```{r}
t(xcol)
```

```{r, error=TRUE}
xcol * xrow
```

```{r, error=TRUE}
xrow * xcol
```

The `%*%` operator works as a matrix multiplication. 

```{r}
xrow %*% xcol
```

## Euclidean norm

To compute the length (norm) of a vector, you can use 

```{r}
sqrt(sum(x ^ 2))
```

The above code works for row and column vectors too.

```{r}
sqrt(sum(xrow ^ 2))
```

## Matrices

Column and row vectors are matrices with a single column and row, respectively. General matrices can be defined similarly.

```{r}
elm_colwise <- c(1.11, 3.21, -5.31, 2.12, -6.22, 0.32)
matrix(elm_colwise, nrow = 3)
```

Notice that by default `matrix()` function fills elements of the passed atomic vector by column. To change this behavior, pass `byrow = TRUE` as parameter.

```{r}
elm_rowwise <- c(
     1.11,  2.12,
     3.21, -6.22,
    -5.31,  0.32
  )
matrix(elm_rowwise, nrow = 3, byrow = TRUE)
```

Addition, subtraction and multiplication are defined for conformal matrices.

```{r}
set.seed(1)
m1 <- matrix(rnorm(9), nrow = 3)
m2 <- matrix(rnorm(9), nrow = 3)
```

Addition:

```{r}
m1 + m2
```

Subtraction:

```{r}
m1 - m2
```

Multiplication: 

```{r}
m1 %*% m2
```

## Comparison

You must not use `==` to check if two non-integer values are identical. 

While you can safely write 

```{r}
1 == 2 - 1
```

you cannot expect that the following comparison returns `TRUE`.

```{r}
0.3 == 0.1 + 0.1 + 0.1
```

When you compare decimal numbers (floating point numbers to be precise),
alway use functions such as `all.equal()`.

```{r}
all.equal(0.3, 0.1 + 0.1 + 0.1)
```

This approach works just fine for matrices and vectors.

```{r}
all.equal(m1, m2) == TRUE
```

# Exercise 1: User defined functions

If you find yourself writing an identical chuck of code over and over again, you should start writing a function. You can reduce the possibility of making bugs by abstracting common tasks because maintenance of codes becomes easier. 

As an exercise, write a user defined function `norm()` to compute Euclidean norm computed above. Ensure that the following code returns the norm:

```{r, eval = FALSE}
norm(x)  #> Should return 4.933559
```

## Solution {-}

```{r}
norm <- function(x) {
  sqrt(sum(x ^ 2))
}
```


# Exercise 2: `source()` external file

Start writing functions in a Rmd file is not always advisable. Since reporting is usually the final stage of your research, you want to start experimentation in a simpler and handier format. The places you might want to write your codes are files with the extension `.R`. 

Look into `R/vector.R` and you can find the `norm()` function defined therein. You can use this function in `R/vector.R` by 

```{r}
source('R/vector.R')
```

and then 

```{r}
norm(x)
```

I recommend to put all `.R` files in the folder named `R` with no subfolders below.^[When you write a R package, you need to follow this rule.]

\pagebreak

# Problems

Edit two files

- `solution.Rmd`
    - Write your name. That's it.
- `matrix.R`
    - Define `is_symmetric()` function that meets the specifications given below. 
    
and knit `solution.Rmd`. If you have a rendered PDF with no warnings, you are done. 
Commit, push and send a Pull Request as always.

## Specifications {-}

Function `is_symmetric()` 

- receives a square matrix `x` as the unique parameter,
- returns `TRUE` for a length-1 vector (scalar),
- returns `TRUE` if `x` is a symmetric matrix and `FALSE` otherwise.

Although you may assume that the parameter `x` is alway square (probably because input validation is done outside of the function), you can make an extra effort to validate that `x` is really square within this function. 

This function is similar to `base::isSymmetric()`.




